---
title: "knn_xwoba"
format: html
editor: visual
---

## Modeling xwOBA with KNN in R

```{r}
#library(tidyverse)

library(tidymodels)
library(kknn)
library(dplyr)
library(yardstick)
library(ggplot2)
```

IMPORTING CLEANED DATA

```{r}
cleaned_data <- readRDS("/Users/chrisa/Documents/GitHub/park-specific-xwoba/data/statcast2024_cleaned.rds")
cleaned_data <- cleaned_data |>
  select(launch_speed,launch_angle,total_bases) |>
  mutate(total_bases = as.factor(total_bases))
```

DEFINING KNN WORKFLOW WITH TIDYMODELS

Model spec

```{r}
#knn_spec <- nearest_neighbor() |>
#  set_mode("classification") |>
#  set_engine(engine = "kknn") |>
#  set_args(neighbors = tune())
```

Variable spec

```{r}
#variable_recipe <- recipe(total_bases ~ launch_speed + launch_angle, data = cleaned_data) |>
#  step_normalize(all_numeric_predictors())
```

Workflow spec (model + recipe)

```{r}
#knn_workflow <- workflow() |>
#  add_recipe(variable_recipe) |>
#  add_model(knn_spec)
```

TUNING K-PARAMETER WITH CROSS-VALIDATION

Using knn_workflow each time, build 200 knn models with 20 levels of k-values.

Evaluate each model using 2-fold cross validation based on accuracy metric.

```{r}
#set.seed(253)
#knn_models <- knn_workflow |>
#  tune_grid(
#    grid = grid_regular(neighbors(range = c(1,200)), levels = 20),
#    resamples = vfold_cv(cleaned_data, v = 2),
#    metrics = metric_set(accuracy)
#  )
```

Comparing knn models and choosing which one to use

```{r}
#knn_models |>
#  collect_metrics()
```

Plotting the different models

```{r}
#knn_models |>
#  autoplot()
```

Finding the best k

```{r}
#best_k <- knn_models |>
#  select_best(metric = "accuracy")
#best_k
```

Getting accuracy for the model with the best k-value

```{r}
#knn_models |>
#  collect_metrics() |>
#  filter(neighbors == best_k$neighbors)
```

FINALIZE AND SAVE FITTED MODEL

Fitting model with the best k-value

```{r}
#final_workflow <- readRDS("/Users/chrisa/Documents/GitHub/park-specific-xwoba/data/best_knn_model.rds")

#final_knn <- knn_workflow |>
#  finalize_workflow(parameters = best_k) |>
#  fit(data = cleaned_data)
```

Save fitted model into 'knn_model_xwoba_saved.'

```{r}
#saveRDS(final_knn, "./data/best_knn_model.rds")
final_knn <- readRDS("/Users/chrisa/Documents/GitHub/park-specific-xwoba/data/best_knn_model.rds")
```

PREDICT TOTAL BASES PROBABILITIES USING FITTED MODEL

Using fitted model to create classification predictions for total_bases based on launch_angle and launch_speed.

```{r}
#cleaned_data_predictions_prob <- final_knn |>
  #predict(new_data = cleaned_data, type = "prob")

#saveRDS(cleaned_data_predictions_prob, "/Users/chrisa/Documents/GitHub/park-specific-xwoba/data/cleaned_data_predictions_prob_knn.rds")

cleaned_data_predictions_prob <- readRDS("/Users/chrisa/Documents/GitHub/park-specific-xwoba/data/cleaned_data_predictions_prob_knn.rds")
```

TESTING ACCURACY

Creating a best prediction column that find the highest probability classification for every event based on the model's predictions.

```{r}
comparison <- cbind(cleaned_data$total_bases, cleaned_data_predictions_prob)
comparison_best_results <- comparison |>
  mutate(best_prediction = case_when(
    .pred_0 == pmax(.pred_0, .pred_1, .pred_2, .pred_3, .pred_4) ~ 0,
    .pred_1 == pmax(.pred_0, .pred_1, .pred_2, .pred_3, .pred_4) ~ 1,
    .pred_2 == pmax(.pred_0, .pred_1, .pred_2, .pred_3, .pred_4) ~ 2,
    .pred_3 == pmax(.pred_0, .pred_1, .pred_2, .pred_3, .pred_4) ~ 3,
    .pred_4 == pmax(.pred_0, .pred_1, .pred_2, .pred_3, .pred_4) ~ 4
  ))
```

Creating a match column where a 1 indicates that the best_prediction of the model and the real value are the same and 0 is where they do not match.

```{r}
accuracy <- comparison_best_results |>
  mutate(match = case_when(
    `cleaned_data$total_bases` == best_prediction ~ 1,
    TRUE ~ 0
  ))
```

Calculate the percentage of total bases correctly predicted.

```{r}
amount_correctly_predicted <- accuracy |>
  count(match == 1)
amount_correctly_predicted_percentage <- (amount_correctly_predicted$n[2] / (amount_correctly_predicted$n[1] + amount_correctly_predicted$n[2])) * 100
amount_correctly_predicted_percentage
```

This accuracy may seem a little low, but the goal of this project is to see how our overall xwOBA values compare to the official values. See the 'Comparing to Official MLB Values' section for more details.

CALCULATING xwOBACON

Weights for 0,1,2,3,4 total bases for 2024 season

```{r}
woba_2024_weights <- c(0, 0.882, 1.254, 1.590, 2.050)
total_bases_probs <- cleaned_data_predictions_prob[c('.pred_0','.pred_1','.pred_2','.pred_3','.pred_4')]
```

```{r}
calculate_xwOBACON <- function(x) {
  return(sum(x * woba_2024_weights))
}
```

```{r}
xwOBACON <- total_bases_probs |>
  mutate(xwOBACON = apply(X = total_bases_probs, MARGIN = 1, FUN = calculate_xwOBACON))
```

RECREATING STATCAST xwOBACON MODEL GRAPH FROM MEDIUM ARTICLE

Recreating "Statcast xwOBACON Model" graph from xwOBA medium article.

Article: <https://medium.com/>@thomasjamesnestico/modelling-xwoba-with-knn-9b004e93861a

```{r}
xwOBACON_graph <- cbind(xwOBACON, cleaned_data$launch_angle, cleaned_data$launch_speed)
```

```{r}
ggplot(xwOBACON_graph, aes(x = `cleaned_data$launch_speed`, y = `cleaned_data$launch_angle`, fill = xwOBACON)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 1)
```

CALCULATING xwOBA

Importing dataset that includes all events, not just ones that were batted into play.

```{r}
final_df <- readRDS("/Users/chrisa/Documents/GitHub/park-specific-xwoba/data/statcast2024_cleaned_all_events.rds")
```

Getting distinct values to be able to left-join the dataset with all events with our calculated xwOBACON values.

```{r}
xwOBACON_w_LA_and_LS <- cbind(xwOBACON, cleaned_data$launch_angle, cleaned_data$launch_speed)

xwOBACON <- xwOBACON_w_LA_and_LS |>
  select(`cleaned_data$launch_speed`, `cleaned_data$launch_angle`, xwOBACON) |>
  distinct()
```

Joining all events dataset with the dataframe where we have calculated xwOBACON. This ensures that the xwOBA values for walks, hit by pitch, strikeout, and strikeout_double_play are correct. This is important because these values play a role in the final calculation of xwOBA.

```{r}
final_df <- final_df |>
  left_join(xwOBACON, by = c("launch_speed" = "cleaned_data$launch_speed",
                             "launch_angle" = "cleaned_data$launch_angle")) |>
  mutate(xwOBA = case_when(
    events == 'walk' ~ 0.689,
    events == 'hit_by_pitch' ~ 0.720,
    events == 'strikeout' ~ 0,
    events == 'strikeout_double_play' ~ 0,
    TRUE ~ xwOBACON))
```

Calculating xwOBA for players with more that 100 plate appearances.

```{r}
cleaned_data_grouped_xwOBA <- final_df |>
  group_by(batter) |>
  #filter(n() > 100) |>
  summarize(xwOBA = mean(xwOBA), woba = mean(woba))
```

COMPARING TO OFFICIAL MLB VALUES

Per Baseball Savant: "\* Qualifiers: 2.1 PA per team game for batters, 1.25 PA per team game for pitchers."

URL: <https://baseballsavant.mlb.com/leaderboard/expected_statistics>

Importing official xwOBA values for players that meet the above qualifications.

```{r}
official_xwOBA_values <- readRDS("/Users/chrisa/Documents/GitHub/park-specific-xwoba/data/mlb_xwoba_2024.rds")
```

Prepping official MLB xwOBA values for inner join with our predicted values.

```{r}
official_xwOBA_values <- official_xwOBA_values |>
  rename(player_id = `xMLBAMID`)

official_xwOBA_values_by_player <- official_xwOBA_values |>
  rename(official_xwoba = xwOBA)

official_xwOBA_values_by_player <- official_xwOBA_values_by_player |>
  select(player_id, official_xwoba)
```

Prepping our predicted xwOBA values for inner join.

```{r}
cleaned_data_grouped_xwOBA_compare <- cleaned_data_grouped_xwOBA |>
  rename(player_id = `batter`)

cleaned_data_grouped_xwOBA_compare <- cleaned_data_grouped_xwOBA_compare |>
  select(player_id, xwOBA)
```

Inner joining the official and our predicted xwOBA values.

```{r}
predicted_vs_official <- official_xwOBA_values_by_player |>
  inner_join(cleaned_data_grouped_xwOBA_compare, join_by(`player_id` == `player_id`))
```

Plotting predicted xwOBA vs official MLB values. Ideally, if we graph the official values on the x-axis and out predicted values on the y-axis, then there should be a perfect 1-to-1 linear fit. Instead, we will fit a linear regression to the graph and see how well our predicted values line up with the official values.

```{r}
fit <- lm(data = predicted_vs_official, official_xwoba ~ xwOBA)
summary(fit)
```

```{r}
predicted_vs_official_plot <- ggplot(data = predicted_vs_official, aes(x = official_xwoba, y = xwOBA)) + 
  geom_point() +
  stat_smooth(method = "lm", col = "red") +
  labs(title = paste("Adj R2 = ",signif(summary(fit)$adj.r.squared, 5),
                     "Intercept =",signif(fit$coef[[1]],5 ),
                     " Slope =",signif(fit$coef[[2]], 5),
                     " P =",signif(summary(fit)$coef[2,4], 5)))
predicted_vs_official_plot
```

The adjusted R\^2 value of 0.92329 indicates that there is a significant linear fit between the official and our predicted values. Considering an R\^2 of 1.0 is a perfect replica of the official values, I would say that our model is a decent open-source version of the xwOBA Statcast model.
