---
title: "knn_xwoba"
format: html
editor: visual
---

## Modeling xwOBA with KNN in R

```{r}
library(tidyverse)
library(tidymodels)
library(rsample)
library(class)
library(kknn)
```

```{r}
cleaned_data <- readRDS("./Data/statcast2024_cleaned.rds")
cleaned_data <- cleaned_data |>
  select('launch_speed','launch_angle','total_bases') |>
  mutate(total_bases = as.factor(total_bases))
```

Model spec

```{r}
knn_spec <- nearest_neighbor() |>
  set_mode("classification") |>
  set_engine(engine = "kknn") |>
  set_args(neighbors = tune())
```

Variable spec

```{r}
variable_recipe <- recipe(total_bases ~ ., data = cleaned_data) |>
  step_normalize(all_numeric_predictors())
```

Workflow spec (model + recipe)

```{r}
knn_workflow <- workflow() |>
  add_recipe(variable_recipe) |>
  add_model(knn_spec)
```

Using knn_workflow each time, build 20 knn models using a range of K tuning parameters from 1 to 40.

Evaluate each model using 2-fold cross validation based on roc_auc metric

```{r}
set.seed(253)
knn_models <- knn_workflow |>
  tune_grid(
    grid = grid_regular(neighbors(range = c(1,40)), levels = 20),
    resamples = vfold_cv(cleaned_data, v = 2),
    metrics = metric_set(roc_auc)
  )
```

Comparing knn models and choosing which one to use

```{r}
knn_models |>
  collect_metrics()
```

Plotting the different models

```{r}
knn_models |>
  autoplot()
```

Finding the best k

```{r}
best_k <- knn_models |>
  select_best(metric = "roc_auc")
best_k
```

Getting accuracy for the model with the best k-value

```{r}
knn_models |>
  collect_metrics() |>
  filter(neighbors == best_k$neighbors)
```

Fitting model with the best k-value

```{r}
final_knn <- knn_workflow |>
  finalize_workflow(parameters = best_k) |>
  fit(data = cleaned_data)
```

Using fitted model to create classification predictions for total_bases based on launch_angle and launch_speed.

```{r}
cleaned_data_predictions_prob <- final_knn |>
  predict(new_data = cleaned_data, type = "prob")
```

Creating a best prediction column that find the highest probability classification for every event based on the model's predictions.

```{r}
comparison <- cbind(cleaned_data$total_bases, cleaned_data_predictions_prob)
comparison_best_results <- comparison |>
  mutate(best_prediction = case_when(
    .pred_0 == pmax(.pred_0, .pred_1, .pred_2, .pred_3, .pred_4) ~ 0,
    .pred_1 == pmax(.pred_0, .pred_1, .pred_2, .pred_3, .pred_4) ~ 1,
    .pred_2 == pmax(.pred_0, .pred_1, .pred_2, .pred_3, .pred_4) ~ 2,
    .pred_3 == pmax(.pred_0, .pred_1, .pred_2, .pred_3, .pred_4) ~ 3,
    .pred_4 == pmax(.pred_0, .pred_1, .pred_2, .pred_3, .pred_4) ~ 4
  ))
```

Creating a match column where a 1 indicates that the best_prediction of the model and the real value are the same and 0 is where they do not match.

```{r}
accuracy <- comparison_best_results |>
  mutate(match = case_when(
    `cleaned_data$total_bases` == best_prediction ~ 1,
    TRUE ~ 0
  ))
accuracy
```
